---
title: "Wrong model, right conclusion: re-analysis of the Ebola virus outbreak in Sierra Leone"
author: Anton Camacho & Joseph Dureau
bibliography: library_papers2.bib
---

```{r global_options, include=FALSE}

dir_repo <- "/Users/Tonton/work/projects/ebola_sierra_leone"

opts_chunk$set(echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE)

```

```{r load_packages, include=FALSE}

library(png)
library(grid)
library(broman)
library(pander)
library(devtools)
library(tidyr)
library(plyr)
library(lubridate)
library(stringr)
library(scales)
library(dplyr)
library(gdata)
library(ggthemr)
ggthemr("fresh")

# tell pander not to split tables
panderOptions('table.split.table', Inf)

```

# Motivation

Modelling work presented in a recent _Nature_ Comment [[@Whitty:2014gc]](http://www.nature.com/news/infectious-disease-tough-choices-to-reduce-ebola-transmission-1.16298) suggests that, without any intervention, the Ebola outbreak in Sierra Leone will peak in early December, reaching a maximum of 1600 weekly reported cases. We argue that this figure is misleading and call for better practice in publishing modelling results.

Assuming that no interventions are in place to reduce transmission, the only reason for the epidemic to peak is via depletion of the susceptible population to a proportion equal $$p_S=\frac{1}{R_0},$$ where $R_0$ is the basic reproduction number (the average number of secondary cases infected by a primary case in a fully susceptible population). Measuring $R_0$ can be difficult because it requires information on the proportion of the population that is immune at the onset of the outbreak. Instead, what is usually measured is the initial reproduction number $$R_{t_0}=R_0\frac{S_{t_0}}{N},$$ that is $R_0$ times the proportion of susceptible (i.e. non-immune) individuals in the population (of size $N$) when the epidemic starts (at time $t_0$). 

However, in the case of a zoonotic virus like Ebola, it is unlikely that a substantial proportion of the Sierra Leone population was immunised by previous exposure to the virus. As such, we can approximate $R_0\approx R_{t_0}$ and $p_S\approx 1/R_{t_0}$. The authors of the _Nature_ Comment report an average reproduction number ($R$, the average number of people infected by each newly infected person) between 1.2 and 1.5 up to early October. This translates into a proportion of susceptible between 67 and 83% at the epidemic peak, that is a cumulative number of cases between `r comma(signif(5743725*17/100,2))` and `r comma(signif(5743725*33/100,2))`.

By contrast, the figure presented in the Comment by [Whitty et al.](http://www.nature.com/news/infectious-disease-tough-choices-to-reduce-ebola-transmission-1.16298) shows approximately `r comma(1600*20/2)` cumulative cases at the epidemic peak (this is a crude upper bound based on a triangular approximation of the incidence curve).

For the current epidemic to peak in December at 1600 weekly reported cases, this would imply that fewer than `r round(100*1600*20/2/(5743725*17/100))`% of infections are reported – a far from credible proportion. Using more realistic assumptions (60% of cases reported), our analysis suggests that without intervention the epidemic will not peak before July 2015, reaching tens of thousand cases per week [2, Figure 1]. In the face of such a nightmare scenario, scaling up the international response as advocated by Whitty et al. appears even more critical. 

The magnitude to which our conclusions differ raises the question of transparency, verifiability and reproducibility of scientific analysis. These are the conditions under which the scientific community will be able to reliably inform decision-making, even more so in urgent situations such as the outbreak we are currently facing. Already, open source technologies are being harnessed to efficiently build and share databases on the current Ebola outbreak. We call for a similar effort to support published modelling results and provide our re-analysis of the data used by Whitty et al. in an open format.


# Data

Weekly number of confirmed plus probable Ebola cases for Sierra-Leone were obtained from the [WHO Situation Report](http://www.who.int/csr/disease/ebola/situation-reports/en/?m=20141112) of 12 November 2014. Since the model presented in Whitty et al. was fitted to data from 13 July to 28 September, we focused on the same time-period.

```{r plot_data, include=TRUE, fig.width=6, fig.height=4}

data <- read.csv(file.path(dir_repo,"SSM/data/data_sierra_leone.csv")) %>% mutate(date=as.Date(date))

p <- ggplot(data, aes(x=date, y=cases_obs__sierra_leone)) + geom_bar(stat="identity") + ylab("Weeky reported cases")
print(p)

```

# Model

We used a SEIR model accounting for delay in notification (cases are reported when they leave the $I_1$ compartment). 

```{r model, fig.width=6, fig.height=2}
img <- readPNG("../figures/ebola_SL.png")
grid.raster(img)
```

The force of infection ($\lambda$) at time $t$ is given by $R_0/D_{inf}(I_1(t)+I_2(t))$.

All durations - incubation period ($D_{inc}$), infectious period ($D_{inf}$), time from onset to notification ($D_{notif}$) - were calibrated using Sierra-Leone specific estimates from the [WHO Ebola Response Team](http://www.nejm.org/doi/abs/10.1056/NEJMoa1411100). In particular, since our model doesn't explicitely account for death and recovery, the infectious period was computed as:

$$ D_{inf} = D_{death}\times CFR + D_{recovery}\times (1-CFR) $$

where $D_{death}$ ($=8.6$ days) and $D_{recovery}$ ($=17.2$ days) are the time from onset to death and to recovery, respectively, and $CFR$ ($=0.69$) is the case fatality ratio. This gives $D_{inf}\approx11$ days.

We estimated the basic reproduction number $R_0$ and the number of infectious individuals at the beginning of the first week in the data (6 July 2014). The observation porcess was modelled as a discretized Normal distribution, whose reporting rate and over-dispersion parameters were also estimated. Let $\tilde Z_t$ be the number of cases reported on the week ending at time $t$, we have the following observation process:

$$ \tilde Z_t \sim  \mathcal{N}(\rho Z_t, \sqrt{\rho (1-\rho) Z_t + (\rho \phi Z_t)^2} )$$

where $Z_t=\int_{t-7}^t I_1(t) dt$ is the true incidence in the model.

We used uniform priors for all parameters except the reporting rate, which was set to an informative Normal prior due to identifiability issues. Indeed, when fitting the increasing phase of an epidemic, the reporting rate is highly correlated to the initial number of infectious and they can't be both identified. Using an informative prior allows us to solve this issue and let some variability in the estimates.


Parameter		Description								Prior	
----------- 	-------------------------------------	------------------- 
$R_0$			Basic reproduction number				$\mathcal{U}(0,5)$	
$D_{inc}$  		Incubation period 						9 days (fixed)	 	
$D_{inf}$		Infectious period						11 days (fixed)		
$D_{notif}$		Time from onset to notification			4.5 days (fixed)		
$\rho$			Reporting rate 							$\mathcal{N}(0.6,0.1)$	
$\phi$			Over-dispersion	in the data 			$\mathcal{U}(0,0.5)$		
$N$				Population size							6,348,450 (fixed)			
$I_1(0)$		Initial number of infectious in $I_1$	$\mathcal{U}(1,200)$		
$I_2(0)$		Initial number of infectious in $I_2$	$I_1(0)*(D_{inf}-D_{notif})/D_{notif}$	
$E(0)$			Initial number of exposed 				$I_1(0)*D_{inc}/D_{notif}$	
$R(0)$			Initial number of removed 				0 (fixed)	
$S(0)$			Initial number of susceptible 			$N - E(0) - I_1(0) - I_2(0) - R(0)$	

# Fit

_In development_

```{r fit, fig.width=10, fig.height=7}
img <- readPNG("../figures/fit_inc_ran_fit_nature_geobs_SL.png")
grid.raster(img)
```

# Forecast

_In development_

```{r forecast, fig.width=10, fig.height=7}
img <- readPNG("../figures/fit_inc_exp_no_improvement.png")
grid.raster(img)
```

# Intervention

_In development_


# References
